---
title: "BIRAX study - Cohort construction"
author: "David M Wright - d.wright@qub.ac.uk"
date: "Document compiled: `r Sys.Date()`"
output: 
  bookdown::html_document2:
     toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(kableExtra)
library(rprojroot)
library(tidyverse)
```

`r R.version.string`   

# Introduction

This analysis is of Electronic Medical Records (EMR) from the Belfast Health and Social Care Trust Macular Service. The aim of the study is to investigate the potential for machine learning to identify factors and biomarkers associated with long-term AMD treatment response.

This document describes construction of the analysis cohort. Pre-processing of the Medisoft and fluid data are described in 'BIRAX-assemble-Medisoft-data-notes.Rmd` and `BIRAX-assemble-NOA-data-notes.Rmd` respectively.

```{r run-analysis, include=FALSE}
source(find_rstudio_root_file("Code/BIRAX-assemble-cohort.R"))
```

 
# Cohort construction

Data were extracted on `r as.Date(max(patients$ExtractDate))`. A proportion of event dates were after the recorded `ExtractDate`. For example, in the `Encounters` table `r nrow(filter(encounters, EncounterDate > ExtractDate))` records had an `EncounterDate` after the `ExtractDate`. This is related to the date perturbation used by Medisoft to maintain patient anonymity in the extract.

The following steps were taken to process the data:   

Injections   

* Removal of a small number of injections that were not for anti-VEGF agents.
* Removal of a small number of duplicate injection entries.   
* Identification of the first injection recorded for each eye and the date (the index date).   
* Calculation of treatment intervals.   

Patient demographics    

* Calculation of patient ages at the index date and at study period end and total observation period for each eye.   

Visual acuity

* Conversion of visual acuity measurements to both ETDRS and logMAR formats.   
* Categorisation of visual acuity into 4 Snellen categories (calculations in logMAR units).   
* Identified baseline VA measurement (closest measurement to index date: earliest date 2 weeks prior, could not occur after index date).   
* At each VA measurement calculate VA change (logMAR and ETDRS) since the baseline (index date).    
* Categorise VA change since baseline in ETDRS lines (calculations in logMAR units)).  
* Identification of the closest VA measurements to each time point of interest during treatment.   

Diagnoses   

* Identification of eyes with a diagnosis of AMD (`CoPathologyCode` "R316" or "R337" in Table \@ref(tab:co-pathology), or DiagnosisDescription containing "AMD" or "age-related macular degeneration" in Table \@ref(tab:diagnoses). Those with "suspected AMD" or "dry AMD" at first diagnosis were excluded (those developing dry AMD later were not excluded)).   
* Identification of eyes with a diagnosis of retinal vein or artery occlusions (RVO).   
* Identification of eyes with a diagnosis of diabetic macular oedema (DMO) or diabetic retinopathy (DR)  (`CoPathologyCode` "R318" in Table \@ref(tab:co-pathology) or `DiagnosisDescription` contining "diabetic macular oedema" in Table \@ref(tab:diagnoses) or any features found in Table \@ref(tab:DR-assessment)).   


OCT thickness metrics

Fluid measurements


## Inclusion and exclusion criteria

Data were extracted by Medisoft with the following inclusion and  exclusion criteria. 

Inclusion: "Any patient having attended the BHSCT macular treatment service at least once during the period May 2008 to May 2018 with a diagnosis of AMD, DR/DMO or RVO."   
Exclusion: "Any patient marked as a test patient."

EMR for a total of `r nrow(patients)` patients were included in the extracted dataset, with `r nrow(injections_raw)` records of anti-VEGF injections. The unit of analysis is the eye so criteria were applied at the eye level:   

* The raw data contained `r nrow(eye_raw)` eyes.   
* `r nrow(filter(eye_raw, exclude_no_AMD))` eyes had no diagnosis of age-related macular degeneration and were excluded. 
* `r nrow(filter(eye_raw, !exclude_no_AMD, exclude_age))` eyes were in patients aged <50 at the index date and were excluded.
* `r nrow(filter(eye_raw, !exclude_no_AMD, !exclude_age,  exclude_lt3_injections))` eyes had fewer than three anti-VEGF injections during the study period  and were excluded.   
* `r nrow(filter(eye_raw, !exclude_no_AMD, !exclude_age,  !exclude_lt3_injections, exclude_RVO))` eyes had a diagnosis of RVO at any point during the study period and were excluded.   
* `r nrow(filter(eye_raw, !exclude_no_AMD, !exclude_age,  !exclude_lt3_injections, !exclude_RVO, exclude_DR_DMO))` eyes had a diagnosis of DMO/DR at any point during the study period and were excluded.   
* `r nrow(filter(eye_raw, !exclude_no_AMD, !exclude_age,  !exclude_lt3_injections, !exclude_RVO, !exclude_DR_DMO, exclude_no_va))` eyes did not have a visual acuity assessment within 2 weeks prior to or including the index date and were excluded.   


A total of `r nrow(eye)` eyes from `r nrow(distinct(eye, PatientID))` patients remained after these exclusions. A total of `r nrow(injections)` injections were observed for the cohort, allowing `r nrow(treatment_intervals)` treatment intervals to be observed (between injections). 


# Descriptive statistics at baseline

Descriptive statistics for the eyes included in the analysis are given in Tables \@ref(tab:descriptives). Statistics are presented at the eye level as some patients were treated in both eyes, with treatment starting at different times. 

Table \@ref(tab:va-descriptives) gives visual acuity outcomes at the times of interest listed in Table \@ref(tab:snapshots). The distribution of OCT assessments, injections and injection intervals are given by treatment year in Tables \@ref(tab:oct-by-year), \@ref(tab:injections-by-year) and \@ref(tab:injection-intervals-by-year).

```{r descriptives}
eye %>%
   transmute(Age = index_age, Gender, 
             `OCT thickness measured` = thickness_measurement, 
             `OCT fluid measured` = fluid_measurement, Eyes = " ") %>%
    # transmute(Age = index_age) %>% 
  GenerateDescriptives(type = c(Age = "Mean (Range)")) %>%
  kbl(caption = "Baseline characteristics of eyes with AMD. Frequency and percentage reported for categorical variables.") %>%
  kable_paper()
```

## Visual acuity at baseline

```{r va-baseline}
va_history %>% 
  filter(baseline) %>% 
  transmute(va_logmar, va_etdrs, va_category_snellen, Eyes = "n") %>% 
  GenerateDescriptives(type = c(va_logmar = "Mean (Range)", va_etdrs = "Median (IQR)")) %>% 
  left_join(var_desc, by = "Variable") %>%
  mutate(Description = coalesce(Description, Variable)) %>%
  select(Description, everything(), -Variable) %>%
  kbl(caption = "Distribution of visual acuity outcomes at baseline.") %>% 
  kable_paper()
```

## Retinal fluid distribution at baseline

These parameters summarised the general scan characteristics in terms of the number of B-scans taken and the number in which fluid were detected. 

```{r fluid-general-baseline}
fluid_history %>% 
  filter(baseline) %>% 
  select(all_of(noa_general)) %>% 
    mutate(Eyes = " ") %>% 
   GenerateDescriptives(type = c(FramesNum = "Median (Range)", NumBscanwtihFluid = "Median (Range)")) %>% 
  left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   kbl(caption = "Distribution of general NOA fluid measurements at baseline") %>% 
   kable_paper()
```

The presence and volume of IRF, SRF and total retinal fluid (TRF) by injection status is given in Table \@ref(tab:fluid-vs-injections).

```{r fluid-baseline}
fluid_history %>% 
  filter(baseline) %>% 
  select(all_of(noa_fluid)) %>% 
    mutate(Eyes = " ") %>% 
   GenerateDescriptives() %>% 
   left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   kbl(caption = "Distribution of NOA OCT fluid measurements at baseline.") %>% 
   kable_paper()
```

Measurements of retinal volume and the presence of vitreo-macula interface or RPE irregularities and detachments by injection status is given in Table \@ref(tab:retinal-vs-injections).

```{r fluid-retinal-baseline, warning=FALSE}
fluid_history %>% 
  filter(baseline) %>% 
  select(all_of(noa_retinal)) %>% 
    mutate(Eyes = " ") %>% 
   GenerateDescriptives() %>% 
   left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   kbl(caption = "Distribution of NOA OCT retinal measurements at baseline.") %>% 
   kable_paper()
```


Measurements of fluid and retinal thickness are given for each of the regions on an ETDRS grid by injection status in Table \@ref(tab:grid-vs-injections).

```{r fluid-grid-baseline, warning=FALSE}
fluid_history %>% 
  filter(baseline) %>% 
  select(all_of(noa_grid)) %>% 
    mutate(Eyes = " ") %>% 
   GenerateDescriptives() %>% 
   left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   kbl(caption = "Distribution of NOA OCT fluid and retinal measurements at baseline by ETDRS grid region.") %>% 
   kable_paper()
```





