---
title: "BIRAX study - NOA analysis Notes"
author: "David M Wright - d.wright@qub.ac.uk"
date: "Document compiled: `r Sys.Date()`"
output: 
  bookdown::html_document2:
     toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(kableExtra)
library(rprojroot)
library(skimr)
```

`r R.version.string`   

# Introduction

This analysis is of performance of the Notal Ophthalmic Analyser (NOA) software to analyse Optical Coherence Tomography (OCT) scans taken of AMD patients at the Belfast Health and Social Care Trust Macular Service. The aim of the study is to compare the NOA outputs regarding AMD lesion activation (i.e. presence of intra-retinal fluid or sub-retinal fluid, IRF and SRF) with the clinical treatment decisions made, as recorded in Electronic Medical Records (EMR).


```{r run-analysis, include=FALSE}
source(find_rstudio_root_file("Code/BIRAX-NOA-analysis.R"))
```


# Pre-processing 

The NOA was run on a heterogeneous collection of OCT scans of various sizes and resolutions, including those of patients with DMO and other exudative retinopathies, over 300,000 scans in total. OCT scans were captured on the Heidelberg Spectralis platform and were extracted and anonymised by the EMR provider (Medisoft) before transfer to the research team. Extracted volume scans were provided in two formats, sets of 2D B-scan images (bmp), and videos stitched together from the component B-scans (AVI). The NOA was run on AVI files. Current progress of the NOA in terms of scans processed is summarised in Table \@ref(tab:noa-logs). Unsuccessful scans were those that could not be processed by the NOA and were discarded. Reasons for unsuccessful processing are given in Table \@ref(tab:noa-unsuccessful). Successful scans were labelled by eligibility for inclusion in the analysis. Ineligible scans were those where the NOA outputs are likely to be unreliable. 

```{r noa-logs}
noa_log %>% 
  kbl(caption = "Progress of the NOA software in processing OCT scans.") %>% 
  kable_paper()
```
```{r noa-unsuccessful}
tabyl(noa_unsuccessful, Details) %>% 
   adorn_totals() %>% 
   adorn_pct_formatting() %>% 
      kbl(caption = "Reasons given for unsuccessful NOA analysis.") %>% 
      kable_paper()
```

## NOA output

The NOA output data structure is defined in the file `NOA_Dictionary.xls` supplied by Notal Vision. Briefly, the presence or absence of IRF and SRF is indicated, along with volume measurements if fluid is detected, in regions defined according to the ETDRS grid.

Bullet points describe operations performed on the data. 

* Patient and eye identifiers extracted from `FileName` column.   
* Where multiple scans were taken for the same eye on the same day, select a single scan by applying the following criteria, breaking ties with the next criterion on the list:   
   + Scan with the fewest missing values, indicating a more successful NOA analysis.   
   + Scan with the largest number of B-scans (`FrameNum`).   
   + Last scan taken on that day.    
   

# Cohort construction

Construction of the main study cohort has been previously described in `BIRAX-analysis-notes-yyyy-mm-dd.html`. It consists of eyes receiving anti-VEGF treatment for neovascular AMD.

For this analysis, the entry date for each cohort member was the date of the first anti-VEGF injection (the index date).
The exit date for each cohort member was the date of the last recorded clinic visit (technically the end interval is left open). Note that this is different to the study period for the injection interval analysis, which only considers injection visits and ends at the date of the last injection.

Each OCT scan was matched to a clinic visit. The primary outcome was whether an injection was administered or not at that visit. This analysis considers the predictive ability of the metrics produced by the NOA in predicting injection, the main metrics of interest being the presence, volume and location of IRF and SRF. 


## Inclusion and exclusion criteria

There were `r nrow(eye)` eyes from `r nrow(distinct(eye, PatientID))` patients in the original study cohort.

* `r nrow(eye) - nrow(eye_oct)` eyes were excluded because there were no matching OCT scans for any visit.

A total of `r nrow(eye_oct)` eyes from `r nrow(distinct(eye_oct, PatientID))` patients remained after these exclusions. There was a total of `r nrow(oct_visits)` visits in which OCT scans were taken.

# Descriptive statistics

Descriptive statistics for the eyes included in the analysis are given in Tables \@ref(tab:descriptives). Statistics are presented at the eye level as some patients were treated in both eyes, with treatment starting at different times. 

```{r descriptives}
eye_oct %>%
  transmute(Age = index_age, Gender, Eyes = " ") %>%
  GenerateDescriptives(type = c(Age = "Mean (Range)")) %>%
  kbl(caption = "Baseline demographics of eyes with AMD. Frequency and percentage reported for categorical variables.") %>%
  kable_paper()
```

# NOA measurements

The distribution of NOA measurements by injection status is given in Tables \@ref(tab:general-vs-injections) to \@ref(tab:grid-vs-injections). Tests of association between each measurement and injection status are presented: t-tests for continuous measurements and chi-squared tests for categorical measurements. 

In general, scans taken at visits at which injections were administered were more likely to have fluid and to have greater fluid volumes than scans taken at visits with no injection.

## General scan characteristics

These parameters summarised the general scan characteristics in terms of the number of B-scans taken and the number in which fluid were detected. 

```{r general-vs-injections}
oct_visits %>% select(injected, all_of(noa_general)) %>% 
    mutate(Total = "") %>% 
   GenerateDescriptives(col_var = injected, output_n = TRUE) %>% 
   left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   left_join(SummariseAssocTest(oct_visits, noa_general, injected), by = "Variable") %>% 
   kbl(caption = "Distribution of general NOA OCT measurements by injection status. Square brackets indicate number of valid measurements.") %>% 
   kable_paper()
```
## Fluid measurements

The presence and volume of IRF, SRF and total retinal fluid (TRF) by injection status is given in Table \@ref(tab:fluid-vs-injections).

```{r fluid-vs-injections}
oct_visits %>% select(injected, noa_fluid) %>% 
    mutate(Total = "") %>% 
   GenerateDescriptives(col_var = injected, output_n = TRUE) %>% 
   left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   left_join(SummariseAssocTest(oct_visits, noa_fluid, injected), by = "Variable") %>% 
   kbl(caption = "Distribution of NOA OCT fluid measurements by injection status. Square brackets indicate number of valid measurements.") %>% 
   kable_paper()
```

## Retinal measurements

Measurements of retinal volume and the presence of vitreo-macula interface or RPE irregularities and detachments by injection status is given in Table \@ref(tab:retinal-vs-injections).

```{r retinal-vs-injections, warning=FALSE}
oct_visits %>% select(injected, all_of(noa_retinal)) %>% 
    mutate(Total = "") %>% 
   GenerateDescriptives(col_var = injected, output_n = TRUE) %>% 
   left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   left_join(SummariseAssocTest(oct_visits, noa_retinal, injected), by = "Variable") %>% 
   kbl(caption = "Distribution of NOA OCT retinal measurements by injection status. Square brackets indicate number of valid measurements.") %>% 
   kable_paper()
```

## ETDRS grid parameters

Measurements of fluid and retinal thickness are given for each of the regions on an ETDRS grid by injection status in Table \@ref(tab:grid-vs-injections).

```{r grid-vs-injections}
oct_visits %>% select(injected, all_of(noa_grid)) %>% 
    mutate(Total = "") %>% 
   GenerateDescriptives(col_var = injected, output_n = TRUE) %>% 
   left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   inner_join(SummariseAssocTest(oct_visits, noa_grid, injected), by = "Variable") %>% 
   arrange(Variable) %>% 
   kbl(caption = "Distribution of NOA OCT retinal measurements by ETDRS grid location and injection status. Square brackets indicate number of valid measurements.") %>% 
   kable_paper()
```


# Diagnostic test characteristics

One potential application of the NOA is to triage OCT scans, classifying those where injections are likely to be necessary and those where no fluid is detected and treatment is not required.

The diagnostic test characteristics of NOA output measures were calculated. For categorical variables (e.g. presence or absence of SRF), accuracy, sensitivity and specificity were calculated. For continuous variables, Receiver Operating Characteristic (ROC) curves were calculated. Variables are ranked by decreasing accuracy or area under the ROC curve.


## General scan characteristics

Of the two general scan characteristics, the number of B-scans with fluid detected had reasonable discrimination (Table \@ref(tab:roc-general)).


```{r roc-general, message=FALSE}
DiagnosticROC(oct_visits, injected, noa_general) %>% 
left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   arrange(desc(AUC)) %>% 
   kbl(caption = "Diagnostic test characteristics of continuous NOA OCT general measurements.", digits = 3) %>% 
   kable_paper()
```

## Fluid measurements

Of the three binary measures of fluid, SRF showed moderate sensitivity (Table \@ref(tab:roc-fluid-binary)). Specificity was good for total fluid. Continuous measures of SRF and TRF had moderate AUC values (Table \@ref(tab:roc-fluid)).

```{r roc-fluid-binary}
noa_diag_stats %>% 
   inner_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`) %>% 
   filter(`Raw Parameter` %in% noa_fluid),
   by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   arrange(desc(accuracy)) %>% 
   kbl(caption = "Diagnostic test characteristics of binary OCT fluid measurements. P value compares accuracy to no information rate.", digits = 3) %>% 
   kable_paper()
```

```{r roc-fluid, message=FALSE}
DiagnosticROC(oct_visits, injected, noa_fluid) %>% 
left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   arrange(desc(AUC)) %>% 
   kbl(caption = "Diagnostic test characteristics of continuous NOA OCT fluid measurements.", digits = 3) %>% 
   kable_paper()
```

## Retinal measurements

Abnormalities at the vitreo-macular interface had high sensitivity and low specificity. Retinal pigment epithelium detachment showed the opposite pattern (Table \@ref(tab:roc-retinal-binary)). Of the continuous retinal measurements, only average retinal height and retinal volume showed moderate AUC values (Table \@ref(tab:roc-retinal)).

```{r roc-retinal-binary}
noa_diag_stats %>% 
   inner_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`) %>% 
   filter(`Raw Parameter` %in% noa_retinal),
   by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   arrange(desc(accuracy)) %>% 
   kbl(caption = "Diagnostic test characteristics of binary OCT retinal measurements. P value compares accuracy to no information rate.", digits = 3) %>% 
   kable_paper()
```

```{r roc-retinal, message=FALSE}
DiagnosticROC(oct_visits, injected, noa_retinal) %>% 
left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   arrange(desc(AUC)) %>% 
   kbl(caption = "Diagnostic test characteristics of continuous NOA OCT retinal measurements.", digits = 3) %>% 
   kable_paper()
```

## ETDRS grid parameters

TRF volumes and sub-field thickness in the central, inner nasal, inner inferior and inner temporal ETDRS regions showed the highest discriminative power, with moderate AUC values (Table \@ref(tab:roc-grid)).


```{r roc-grid, message=FALSE}
DiagnosticROC(oct_visits, injected, noa_grid) %>% 
left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   arrange(desc(AUC)) %>% 
   kbl(caption = "Diagnostic test characteristics of continuous NOA OCT measurements by ETDRS region.", digits = 3) %>% 
   kable_paper()
```

