---
title: "BIRAX study - Cohort construction"
author: "David M Wright - d.wright@qub.ac.uk"
date: "Document compiled: `r Sys.Date()`"
output: 
  bookdown::html_document2:
     toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(kableExtra)
library(rprojroot)
library(tidyverse)
library(broom)
```

`r R.version.string`   

# Introduction

```{r run-analysis, include=FALSE}
source(find_rstudio_root_file("Code/BIRAX-assemble-cohort.R"))
```

This analysis is of Electronic Medical Records (EMR) from the Belfast Health and Social Care Trust Macular Service extracted over the period `r min(as.Date(encounters$EncounterDate))` to `r max(as.Date(encounters$ExtractDate))`. The aim of the study is to investigate the potential for machine learning to identify factors and biomarkers associated with long-term AMD treatment response.

This document describes construction of the analysis cohort. Pre-processing of the Medisoft and fluid data are described in `BIRAX-assemble-Medisoft-data.html` and `BIRAX-assemble-NOA-data.html` respectively.

 
# Cohort construction

Data were extracted on `r as.Date(max(patients$ExtractDate))`. A proportion of event dates were after the recorded `ExtractDate`. For example, in the `Encounters` table `r nrow(filter(encounters, EncounterDate > ExtractDate))` records had an `EncounterDate` after the `ExtractDate`. This is related to the date perturbation used by Medisoft to maintain patient anonymity in the extract.

The following steps were taken to process the data:   

Injections   

* Removal of a small number of injections that were not for anti-VEGF agents.
* Removal of a small number of duplicate injection entries.   
* Identification of the first injection recorded for each eye and the date (the index date).   
* Calculation of treatment intervals.   
* Calculate the total time under treatment.

Patient demographics    

* Calculation of patient ages at the index date and at study period end and total observation period for each eye (index age to age at death or study period end, whichever was sooner).    
* In ~10% of cases the total time under treatment was greater than the calculated observation period by up to half a year. Where this occurred set the total observation period equal to the total time under treatment.   

Visual acuity

* Conversion of visual acuity measurements to both ETDRS and logMAR formats.   
* Categorisation of visual acuity into 4 Snellen categories (calculations in logMAR units).   
* Identified baseline VA measurement (closest measurement to index date: earliest date 2 weeks prior, could not occur after index date).   
* At each VA measurement calculate VA change (logMAR and ETDRS) since the baseline (index date).    
* Categorise VA change since baseline in ETDRS lines (calculations in logMAR units)).  
* Identification of the closest VA measurement to each time point of interest during treatment.   

Diagnoses   

* Identification of eyes with a diagnosis of AMD (`CoPathologyCode` "R316" or "R337" in `BELCoPathology` table, or `DiagnosisDescription` containing "AMD" or "age-related macular degeneration" in `BELDiagnoses` table). Those with "suspected AMD" or "dry AMD" at first diagnosis were excluded (those developing dry AMD later were not excluded)).   
* Identification of eyes with a diagnosis of retinal vein or artery occlusions (RVO).   
* Identification of eyes with a diagnosis of diabetic macular oedema (DMO) or diabetic retinopathy (DR)  (`CoPathologyCode` "R318" in `BELCoPathology` table or `DiagnosisDescription` containing "diabetic macular oedema" in `BELDiagnoses` table or any features found in `BELDRAssessment` table).   


OCT thickness metrics

These metrics were derived from the standard Heidelberg segmentation and give retinal thickness and volume by region of the ETDRS grid.

* Identified baseline OCT thickness measurement (closest measurement to index date: earliest date 2 weeks prior, could not occur after index date).   
* Identification of the closest OCT measurement to each time point of interest during treatment.    

Fluid measurements

* Identified baseline fluid measurement (closest measurement to index date: earliest date 2 weeks prior, could not occur after index date).   
* Identification of the closest fluid measurement to each time point of interest during treatment.    
* Categorised AMD type at each fluid measurement: None/SRF only/IRF only/Both SRF and IRF.   


## Inclusion and exclusion criteria

Data were extracted by Medisoft with the following inclusion and  exclusion criteria. 

Inclusion: "Any patient having attended the BHSCT macular treatment service at least once during the period May 2008 to May 2018 with a diagnosis of AMD, DR/DMO or RVO."   
Exclusion: "Any patient marked as a test patient."

EMR for a total of `r nrow(patients)` patients were included in the extracted dataset, with `r nrow(injections_raw)` records of anti-VEGF injections. The unit of analysis is the eye so criteria were applied at the eye level:   

* The raw data contained `r nrow(eye_raw)` eyes.   
* `r nrow(filter(eye_raw, exclude_no_AMD))` eyes had no diagnosis of age-related macular degeneration and were excluded. 
* `r nrow(filter(eye_raw, !exclude_no_AMD, exclude_age))` eyes were in patients aged <50 at the index date and were excluded.
* `r nrow(filter(eye_raw, !exclude_no_AMD, !exclude_age, exclude_lt3_injections))` eyes had fewer than three anti-VEGF injections during the study period  and were excluded.   
* `r nrow(filter(eye_raw, !exclude_no_AMD, !exclude_age, !exclude_lt3_injections, exclude_RVO))` eyes had a diagnosis of RVO at any point during the study period and were excluded.   
* `r nrow(filter(eye_raw, !exclude_no_AMD, !exclude_age,  !exclude_lt3_injections, !exclude_RVO, exclude_DR_DMO))` eyes had a diagnosis of DMO/DR at any point during the study period and were excluded.   
* `r nrow(filter(eye_raw, !exclude_no_AMD, !exclude_age,  !exclude_lt3_injections, !exclude_RVO, !exclude_DR_DMO, exclude_no_va))` eyes did not have a visual acuity assessment within 2 weeks prior to or including the index date and were excluded.   
* `r nrow(filter(eye_raw, !exclude_no_AMD, !exclude_age,  !exclude_lt3_injections, !exclude_RVO, !exclude_DR_DMO, !exclude_no_va, exclude_no_thickness))` eyes did not have an OCT retinal thickness measurement within 2 weeks prior to or including the index date and were excluded.   
* `r nrow(filter(eye_raw, !exclude_no_AMD, !exclude_age,  !exclude_lt3_injections, !exclude_RVO, !exclude_DR_DMO, !exclude_no_va, !exclude_no_thickness, exclude_no_fluid))` eyes did not have a retinal fluid measurement within 2 weeks prior to or including the index date and were excluded.   


A total of `r nrow(eye)` eyes from `r nrow(distinct(eye, PatientID))` patients remained after these exclusions. A total of `r nrow(injections)` injections were observed for the cohort, allowing `r nrow(treatment_intervals)` treatment intervals to be observed (between injections). 

# Descriptive statistics at baseline

Descriptive statistics for the eyes included in the analysis are given in Tables \@ref(tab:descriptives) to \@ref(tab:fluid-grid-baseline). Statistics are presented at the eye level as some patients were treated in both eyes, with treatment starting at different times. 

## Patient characteristics

```{r descriptives}
eye %>%
   transmute(Age = index_age, Gender, Eyes = " ") %>%
  GenerateDescriptives(type = c(Age = "Mean (Range)")) %>%
  kbl(caption = "Baseline characteristics of eyes with AMD. Frequency and percentage reported for categorical variables.") %>%
  kable_paper()
```
A key characteristic of the dataset is the different lengths of follow-up overall (i.e. time period for which records were extracted), in terms of treatment duration and in terms of monitoring for each examination type, given in Table \@ref(tab:follow-up-length). There are several different measures presented:

* _Years observed_ gives the time from the first injection (baseline) to the date of death (if death occurred) or the date at which the data were extracted for the study.   
* _Years treated_ gives the time from the first injection to the last recorded injection.   
* _VA years observed_ gives the time from the first injection to the last recorded VA measurement.   

An eye may have a much longer _Years observed_ than _Years treated_ or _VA years observed_ period. This may occur if a patient is discharged from the macular service when treatment ceases. This may be due to onset of GA or build up of fibrous tissue, such that continued anti-VEGF treatment is unlikely to be beneficial. A patient may have VA monitored for an additional year in the macular service after treatment cessation but this might also occur in a low-vision clinic, which is likely to be more convenient for the patient. 

```{r follow-up-length}
bind_rows(
eye %>% 
  transmute(Eyes = " ",
            `Years observed` = years_observed,
            `Total injections` = total_injections,
            Injections = total_injections,
            `Years treated` = years_treated
            ) %>% 
  GenerateDescriptives(type = c(`Years observed` = "Median (Range)", 
                                Injections = "Median (Range)",
                                `Years treated` = "Median (Range)",
                                `Total injections` = " ")),

GenerateDescriptives(transmute(va_history, `Total VA exams` = " ")),

va_history %>% 
  group_by(PatientID, EyeCode) %>% 
  summarise(`VA exams` = n(),
            `VA years observed` = max(months_since_index/12), .groups = "drop") %>% 
  select(-PatientID, -EyeCode) %>% 
    GenerateDescriptives(type = c(`VA exams` = "Median (Range)",
      `VA years observed` = "Median (Range)")),

va_history %>% 
  group_by(PatientID, EyeCode) %>% 
  summarise(`VA years observed` = max(months_since_index/12), .groups = "drop") %>% 
  select(-PatientID, -EyeCode) %>% 
    GenerateDescriptives(),

GenerateDescriptives(transmute(thickness_history, `Total OCT thickness measurements` = " ")),

thickness_history %>% 
  group_by(PatientID, EyeCode) %>% 
  summarise(`OCT thickness measurements` = n(),
            `OCT thickness years observed` = max(months_since_index/12), .groups = "drop") %>% 
  select(-PatientID, -EyeCode) %>% 
    GenerateDescriptives(type = c(`OCT thickness measurements` = "Median (Range)",
      `OCT thickness years observed` = "Median (Range)")),

GenerateDescriptives(transmute(fluid_history, `Total OCT fluid measurements` = " ")),

fluid_history %>% 
  group_by(PatientID, EyeCode) %>% 
  summarise(`OCT fluid measurements` = n(),
            `OCT fluid years observed` = max(months_since_index/12), .groups = "drop") %>% 
  select(-PatientID, -EyeCode) %>% 
    GenerateDescriptives(type = c(`OCT fluid measurements` = "Median (Range)",
      `OCT fluid years observed` = "Median (Range)"))
) %>%
  kbl(caption = "Number of examinations and length of follow-up.") %>% 
  kable_paper()
```

## Follow-up by treatment year

```{r follow-up-by-year}
injections %>% 
  count(PatientID, EyeCode, treatment_year, name = "Injections") %>% 
  select(treatment_year, Injections) %>% 
  GenerateDescriptives(col_var = treatment_year) %>% 
  bind_rows(
injections %>% 
  count(PatientID, EyeCode, treatment_year, name = "Injections") %>% 
  transmute(treatment_year, Injections, Eyes = " ") %>% 
  GenerateDescriptives(col_var = treatment_year, type = c(Injections = "Median (Range)"))
  ) %>% 
  kbl(caption = "Distribution of injections received by year of anti-VEGF treatment") %>% 
  kable_paper()
```

```{r follow-up-by-year-amd-type}
injections %>% 
  inner_join(
  fluid_history %>% 
  filter(baseline) %>% 
  select(PatientID, EyeCode, amd_type)
  , by = c("PatientID", "EyeCode")) %>% 
  group_by(amd_type) %>% 
  count(amd_type, PatientID, EyeCode, treatment_year, name = "Injections") %>% 
  select(amd_type, treatment_year, Injections) %>% 
    group_map(.f = ~bind_cols(amd_type = .y, GenerateDescriptives(.x, col_var = treatment_year) )) %>% 
  bind_rows() %>% 
  bind_rows(injections %>% 
  inner_join(
  fluid_history %>% 
  filter(baseline) %>% 
  select(PatientID, EyeCode, amd_type)
  , by = c("PatientID", "EyeCode")) %>% 
  group_by(amd_type) %>% 
  count(amd_type, PatientID, EyeCode, treatment_year, name = "Injections") %>% 
  transmute(amd_type, treatment_year, Injections, Eyes = " ") %>% 
    group_map(.f = ~bind_cols(amd_type = .y, GenerateDescriptives(.x, col_var = treatment_year, type = c(Injections = "Median (Range)") ))) %>% 
  bind_rows()
  ) %>% 
  arrange(amd_type, Variable) %>% 
  kbl(caption = "Distribution of injections received by year of anti-VEGF treatment and fluid type at baseline") %>% 
  kable_paper()
```

```{r va-followup-by-year}
va_history %>% 
  filter(treatment_year != "0") %>% 
  count(PatientID, EyeCode, treatment_year, name = "VA measurements") %>% 
  select(treatment_year, `VA measurements`) %>% 
  GenerateDescriptives(col_var = treatment_year) %>% 
  bind_rows(va_history %>% 
  filter(treatment_year != "0") %>% 
  count(PatientID, EyeCode, treatment_year, name = "VA measurements") %>% 
  transmute(treatment_year, `VA measurements`, Eyes = " ") %>% 
  GenerateDescriptives(col_var = treatment_year, type = c("VA measurements" = "Median (Range)"))) %>% 
  kbl(caption = "Distribution of VA measurements by year of anti-VEGF treatment") %>% 
  kable_paper()
```

```{r va-followup-by-year-amd-type}
va_history %>% inner_join(
  fluid_history %>% 
  filter(baseline) %>% 
  select(PatientID, EyeCode, amd_type)
  , by = c("PatientID", "EyeCode")) %>% 
  group_by(amd_type) %>% 
  filter(treatment_year != "0") %>% 
  count(amd_type, PatientID, EyeCode, treatment_year, name = "VA measurements") %>% 
  select(amd_type, treatment_year, `VA measurements`) %>% 
  group_map(~bind_cols(amd_type = .y, GenerateDescriptives(.x, col_var = treatment_year))) %>% 
              bind_rows() %>% 
  bind_rows(va_history %>% inner_join(
  fluid_history %>% 
  filter(baseline) %>% 
  select(PatientID, EyeCode, amd_type)
  , by = c("PatientID", "EyeCode")) %>% 
  group_by(amd_type) %>% 
  filter(treatment_year != "0") %>% 
  count(amd_type, PatientID, EyeCode, treatment_year, name = "VA measurements") %>% 
  transmute(amd_type, treatment_year, `VA measurements`, Eyes = " ") %>% 
  group_map(~bind_cols(amd_type = .y, GenerateDescriptives(.x, col_var = treatment_year, type = c("VA measurements" = "Median (Range)")))) %>% 
    bind_rows()) %>% 
    arrange(amd_type, desc(Variable)) %>% 
  kbl(caption = "Distribution of VA measurements by year of anti-VEGF treatment") %>% 
  kable_paper()
```

  
## Visual acuity

```{r va-baseline}
va_history %>% 
  filter(baseline) %>% 
  transmute(va_logmar, va_etdrs, va_category_snellen, Eyes = "n") %>% 
  GenerateDescriptives(type = c(va_logmar = "Median (IQR)", va_etdrs = "Median (IQR)")) %>% 
  left_join(var_desc, by = "Variable") %>%
  mutate(Description = coalesce(Description, Variable)) %>%
  select(Description, everything(), -Variable) %>%
  kbl(caption = "Distribution of visual acuity outcomes at baseline.") %>% 
  kable_paper()
```

```{r va-baseline-amd-type}
fluid_history %>% 
  filter(baseline) %>% 
  select(PatientID, EyeCode, amd_type) %>% 
  inner_join(
va_history %>% 
  filter(baseline),
by = c("PatientID", "EyeCode")) %>% 
  transmute(amd_type, va_logmar, va_etdrs, va_category_snellen, Eyes = "n") %>% 
  GenerateDescriptives(col_var = amd_type, type = c(va_logmar = "Median (IQR)", va_etdrs = "Median (IQR)")) %>% 
  left_join(var_desc, by = "Variable") %>%
  mutate(Description = coalesce(Description, Variable)) %>%
  select(Description, everything(), -Variable) %>%
  kbl(caption = "Distribution of visual acuity outcomes at baseline by fluid presence.") %>% 
  kable_paper()
```

## Retinal fluid distribution

General scan characteristics in terms of the number of B-scans taken and the number in which fluid were detected is given in Table \@ref(tab:fluid-general-baseline). The presence and volume of intra-retinal fluid (IRF), sub-retinal fluid (SRF) and total retinal fluid (TRF) is given in Table \@ref(tab:fluid-baseline), along with estimated correlations between each of the continuous measurements and visual acuity (logMAR). Measurements of retinal volume and the presence of vitreo-macula interface or retinal pigmentary epithelium (RPE) irregularities and detachments is given in Table \@ref(tab:fluid-retinal-baseline). Measurements of fluid and retinal thickness are given for each of the regions on an ETDRS grid are given in Table \@ref(tab:fluid-grid-baseline).

```{r fluid-general-baseline}
fluid_history %>% 
  filter(baseline) %>% 
  select(all_of(noa_general)) %>% 
    mutate(Eyes = " ") %>% 
   GenerateDescriptives(type = c(FramesNum = "Median (Range)", NumBscanwtihFluid = "Median (Range)")) %>% 
  left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   kbl(caption = "Distribution of general NOA fluid measurements at baseline") %>% 
   kable_paper()
```

```{r fluid-general-baseline-by-age}
fluid_history %>% 
  filter(baseline) %>% 
  inner_join(eye, by = c("PatientID", "EyeCode")) %>% 
  select(Age, all_of(noa_general)) %>% 
    mutate(Eyes = " ") %>% 
   GenerateDescriptives(col_var = Age, type = c(FramesNum = "Median (Range)", 
                                                NumBscanwtihFluid = "Median (Range)"
                                                )) %>% 
  left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   kbl(caption = "Distribution of general NOA fluid measurements at baseline by age category.") %>% 
   kable_paper()
```

```{r fluid-general-baseline-by-gender}
fluid_history %>% 
  filter(baseline) %>% 
  inner_join(eye, by = c("PatientID", "EyeCode")) %>% 
  select(Gender, all_of(noa_general)) %>% 
    mutate(Eyes = " ") %>% 
   GenerateDescriptives(col_var = Gender, type = c(FramesNum = "Median (Range)", NumBscanwtihFluid = "Median (Range)")) %>% 
  left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   kbl(caption = "Distribution of general NOA fluid measurements at baseline by gender.") %>% 
   kable_paper()
```

```{r fluid-baseline, warning=FALSE}
fluid_history %>% 
  filter(baseline) %>% 
  select(all_of(noa_fluid), `PED volume[nl]`) %>% 
    mutate(Eyes = " ") %>% 
   GenerateDescriptives(type = c(SRFVolumeNl = "Median (IQR)",
                          IRFVolumeNl = "Median (IQR)",
                          TRFVolumeNl = "Median (IQR)",
                          `Average_SRF_height[um]` = "Median (IQR)",
                          `Average_IRF_height[um]` = "Median (IQR)",
                          `Average_TRF_height[um]` = "Median (IQR)",
                          `MaxSRF_height[um]` = "Median (IQR)",
                          `MaxIRF_height[um]` = "Median (IQR)",
                          `PED volume[nl]` = "Median (IQR)")) %>% 
  left_join(va_fl_correlations, by = "Variable") %>% 
   left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   arrange(desc(estimate)) %>% 
   select(-estimate) %>% 
   kbl(caption = "Distribution of NOA OCT fluid measurements at baseline, sorted by strength of correlation with visual acuity.") %>% 
   kable_paper()
```

```{r fluid-baseline-by-age, warning=FALSE}
fluid_history %>% 
  filter(baseline) %>% 
  inner_join(eye, by = c("PatientID", "EyeCode")) %>% 
  select(Age, all_of(noa_fluid), `PED volume[nl]`) %>% 
    mutate(Eyes = " ") %>% 
   GenerateDescriptives(col_var = Age,
                        type = c(SRFVolumeNl = "Median (IQR)",
                          IRFVolumeNl = "Median (IQR)",
                          TRFVolumeNl = "Median (IQR)",
                          `Average_SRF_height[um]` = "Median (IQR)",
                          `Average_IRF_height[um]` = "Median (IQR)",
                          `Average_TRF_height[um]` = "Median (IQR)",
                          `MaxSRF_height[um]` = "Median (IQR)",
                          `MaxIRF_height[um]` = "Median (IQR)",
                          `PED volume[nl]` = "Median (IQR)"
                        )) %>% 
   left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
  
  # Fit a linear trend with age as a continuous variable
  left_join(fluid_history %>%
  filter(baseline) %>%
  inner_join(eye, by = c("PatientID", "EyeCode")) %>%
  select(index_age, all_of(noa_fluid), `PED volume[nl]`) %>%
  map_dfr( .y = .$index_age, ~tidy(lm(.x~.y)), .id = "Variable") %>% 
  filter(term == ".y", !is.na(p.value)) %>% 
  transmute(Variable, Trend = estimate, P = format.pval(p.value, digits = 1,  eps = 0.001)),
  by = "Variable") %>% 
  
  # Assemble the output table
   relocate(Parameter, .after = Variable) %>% 
   kbl(digits = 2, caption = "Distribution of NOA OCT fluid measurements at baseline by age category. P values are for linear trend in age (treated as continuous variable).") %>% 
   kable_paper()

```

```{r fluid-baseline-by-gender, warning=FALSE}
fluid_history %>% 
  filter(baseline) %>% 
  inner_join(eye, by = c("PatientID", "EyeCode")) %>% 
  select(Gender, all_of(noa_fluid), `PED volume[nl]`) %>% 
    mutate(Eyes = " ") %>% 
   GenerateDescriptives(col_var = Gender,
                        type = c(SRFVolumeNl = "Median (IQR)",
                          IRFVolumeNl = "Median (IQR)",
                          TRFVolumeNl = "Median (IQR)",
                          `Average_SRF_height[um]` = "Median (IQR)",
                          `Average_IRF_height[um]` = "Median (IQR)",
                          `Average_TRF_height[um]` = "Median (IQR)",
                          `MaxSRF_height[um]` = "Median (IQR)",
                          `MaxIRF_height[um]` = "Median (IQR)",
                          `PED volume[nl]` = "Median (IQR)"
                        )) %>% 
   left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
  
    
  # Fit a linear trend with age as a continuous variable
  left_join(fluid_history %>%
  filter(baseline) %>%
  inner_join(eye, by = c("PatientID", "EyeCode")) %>%
  select(Gender, all_of(noa_fluid), `PED volume[nl]`) %>%
  map_dfr( .y = .$Gender, ~tidy(lm(.x~.y)), .id = "Variable") %>% 
  filter(term == ".yM", !is.na(p.value)) %>% 
  transmute(Variable, `Difference in means (M-F)` = estimate, P = format.pval(p.value, digits = 1,  eps = 0.001)),
  by = "Variable") %>% 
  
   relocate(Parameter, .after = Variable) %>% 
   kbl(digits = 2, caption = "Distribution of NOA OCT fluid measurements at baseline by gender.") %>% 
   kable_paper()
```


```{r fluid-retinal-baseline, warning=FALSE}
fluid_history %>% 
  filter(baseline) %>% 
  select(all_of(noa_retinal)) %>% 
    mutate(Eyes = " ") %>% 
   GenerateDescriptives(type = c(
     `Average_Retinal_height[um]` = "Median (IQR)",
     Retinal_volumeNl = "Median (IQR)",
     `Maximal_height_of_RPE_irregularities[um]` = "Median (IQR)",
       `PED volume[nl]` = "Median (IQR)",
     `Volume_of_RPE_irregularities[nl]` = "Median (IQR)",
       `AnalyzedEnfaceArea[mm^2]` = "Median (IQR)",
     `Average_RPE_irregularities_height[um]` = "Median (IQR)",
     ERMVolumeNl = "Median (IQR)"
   )) %>% 
   left_join(va_fl_correlations, by = "Variable") %>% 
   left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
     arrange(desc(estimate)) %>% 
   select(-estimate) %>% 
   kbl(caption = "Distribution of NOA OCT retinal measurements at baseline, sorted by strength of correlation with visual acuity.") %>% 
   kable_paper()
```

```{r fluid-retinal-baseline-by-age}
fluid_history %>% 
  filter(baseline) %>% 
  inner_join(eye, by = c("PatientID", "EyeCode")) %>% 
  select(Age, all_of(noa_retinal)) %>% 
    mutate(Eyes = " ") %>% 
   GenerateDescriptives(col_var = Age) %>% 
   left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   kbl(caption = "Distribution of NOA OCT retinal measurements at baseline by age category.") %>% 
   kable_paper()
```

```{r fluid-retinal-baseline-by-gender}
fluid_history %>% 
  filter(baseline) %>% 
  inner_join(eye, by = c("PatientID", "EyeCode")) %>% 
  select(Gender, all_of(noa_retinal)) %>% 
    mutate(Eyes = " ") %>% 
   GenerateDescriptives(col_var = Gender) %>% 
   left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   kbl(caption = "Distribution of NOA OCT retinal measurements at baseline by gender.") %>% 
   kable_paper()
```

```{r fluid-grid-plot, fig.cap = "Distribution of retinal fluid at baseline by ETDRS grid region: mean (standard deviation). Thicknesses measured in um, volumes in nl.", fig.height=12, fig.width=8}
baseline_fluid_grid %>% 
  filter(str_detect(abbr_metric, "SRF|IRF|TRF")) %>% 
  PlotETDRSGrid(valform) +
  facet_wrap(~abbr_metric, ncol = 2)
```
```{r RPE-grid-plot, fig.cap = "Distribution of RPE irregularities at baseline by ETDRS grid region: mean (standard deviation). Thicknesses measured in um, volumes in nl.", fig.height=4, fig.width=8}
baseline_fluid_grid %>% 
  filter(str_detect(abbr_metric, "RPE")) %>% 
  PlotETDRSGrid(valform) +
  facet_wrap(~abbr_metric, ncol = 2)
```

```{r retina-grid-plot, fig.cap = "Distribution of retinal thickness at baseline by ETDRS grid region: mean (standard deviation). Thicknesses measured in um, volumes in nl.", fig.height=4, fig.width=8}
baseline_fluid_grid %>% 
  filter(str_detect(abbr_metric, "retina")) %>% 
  PlotETDRSGrid(valform) +
  facet_wrap(~abbr_metric, ncol = 2)
```

```{r fluid-grid-baseline, warning=FALSE}
fluid_history %>% 
  filter(baseline) %>% 
  select(all_of(noa_grid)) %>% 
    mutate(Eyes = " ") %>% 
   GenerateDescriptives(type = set_names(rep("Median (IQR)", length(noa_grid)), nm = noa_grid)
) %>% 
   left_join(va_fl_correlations, by = "Variable") %>% 
   left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
    arrange(desc(estimate)) %>% 
   select(-estimate) %>% 
   kbl(caption = "Distribution of NOA OCT fluid and retinal measurements at baseline by ETDRS grid region, sorted by strength of correlation with visual acuity.") %>% 
   kable_paper()
```


### Eyes with SRF and IRF at baseline

```{r fluid-baseline-srf-irf, warning=FALSE}
fluid_history %>% 
  filter(baseline, amd_type == "SRF and IRF") %>% 
  select(all_of(noa_fluid), `PED volume[nl]`) %>% 
    mutate(Eyes = " ") %>% 
   GenerateDescriptives(type = c(SRFVolumeNl = "Median (IQR)",
                          IRFVolumeNl = "Median (IQR)",
                          TRFVolumeNl = "Median (IQR)",
                          `Average_SRF_height[um]` = "Median (IQR)",
                          `Average_IRF_height[um]` = "Median (IQR)",
                          `Average_TRF_height[um]` = "Median (IQR)",
                          `MaxSRF_height[um]` = "Median (IQR)",
                          `MaxIRF_height[um]` = "Median (IQR)",
                          `PED volume[nl]` = "Median (IQR)")) %>% 
  left_join(va_fl_correlations_srf_irf, by = "Variable") %>% 
   left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
   arrange(desc(estimate)) %>% 
   select(-estimate) %>% 
   kbl(caption = "Distribution of NOA OCT fluid measurements at baseline, sorted by strength of correlation with visual acuity. Eyes with SRF and IRF at baseline.") %>% 
   kable_paper()
```


```{r fluid-grid-baseline-srf-irf, warning=FALSE}
fluid_history %>% 
  filter(baseline, amd_type == "SRF and IRF") %>% 
  select(all_of(noa_grid)) %>% 
    mutate(Eyes = " ") %>% 
   GenerateDescriptives(type = set_names(rep("Median (IQR)", length(noa_grid)), nm = noa_grid)
) %>% 
   left_join(va_fl_correlations_srf_irf, by = "Variable") %>% 
   left_join(noa_dictionary %>% 
                 select(Parameter, `Raw Parameter`), 
              by = c("Variable" = "Raw Parameter")) %>% 
   relocate(Parameter, .after = Variable) %>% 
    arrange(desc(estimate)) %>% 
   select(-estimate) %>% 
   kbl(caption = "Distribution of NOA OCT fluid and retinal measurements at baseline by ETDRS grid region, sorted by strength of correlation with visual acuity. Eyes with SRF and IRF at baseline.") %>% 
   kable_paper()
```