---
title: "Diabetes study - Cohort construction"
author: "David M Wright - d.wright@qub.ac.uk"
date: "Document compiled: `r Sys.Date()`"
output: 
  bookdown::html_document2:
     toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(kableExtra)
library(rprojroot)
library(tidyverse)
library(broom)
```

`r R.version.string`   

# Introduction

```{r run-analysis, include=FALSE}
source(find_rstudio_root_file("Code/DM-cohort-construction.R"))
```

This analysis is of Electronic Medical Records (EMR) from the Belfast Health and Social Care Trust Macular Service extracted over the period `r min(as.Date(encounters$EncounterDate))` to `r max(as.Date(encounters$ExtractDate))`. The aim of the study is to investigate treatment outcomes among patients with Diabetes Mellitus (DM).

This document describes construction of the analysis cohort. Pre-processing of the Medisoft data is described in `BIRAX-assemble-Medisoft-data.html`.

 
# Cohort construction

Data were extracted on `r as.Date(max(patients$ExtractDate))`. A proportion of event dates were after the recorded `ExtractDate`. For example, in the `Encounters` table `r nrow(filter(encounters, EncounterDate > ExtractDate))` records had an `EncounterDate` after the `ExtractDate`. This is related to the date perturbation used by Medisoft to maintain patient anonymity in the extract.

The following steps were taken to process the data.   

Identification of patients with diabetes:

* Patient was recorded in the `BELDiabeticDiagnosis` table   
* Or had diabetic retinopathy recorded:   
  + `CoPathologyCode` R317 in `BELCoPathology` table   
  + or `DiagnosisDescription` containing "diabetic retinopathy" in `BELDiagnoses`   
  + or `ClinicalTermDescription` containing "diabetic retinopathy" in `BELClinicalFindings`   
  + or `DRAsessmentDesc` not containing "DMO, oedema, CSMO or none" in `BELDRAssessment`   
* Or had diabetic macular oedema recorded:   
  + or `DiagnosisDescription` containing "diabetic macular oedema" in `BELDiagnoses`   
  + or `ClinicalTermDescription` containing "diabetic macular oedema" in `BELClinicalFindings`   
  + or `DRAsessmentDesc` containing "DMO, oedema, CSMO" in `BELDRAssessment`


## Inclusion and exclusion criteria

Data were extracted by Medisoft with the following inclusion and  exclusion criteria. 

Inclusion: "Any patient having attended the BHSCT macular treatment service at least once during the period May 2008 to May 2018 with a diagnosis of AMD, DR/DMO or RVO."   
Exclusion: "Any patient marked as a test patient."

EMR for a total of `r nrow(patients)` patients were included in the extracted dataset. The unit of analysis is the eye so criteria were applied at the eye level:   

* The raw data potentially contained `r nrow(eye_raw_dm)` eyes (some were not noted in any of the tables and so assumed normal).   
* `r nrow(filter(eye_raw_dm, exclude_no_dm))` eyes had no diagnosis of DM and were excluded.    

This left `r nrow(filter(eye_raw_dm, !exclude_no_dm))` DM eyes from `r nrow(distinct(filter(eye_raw_dm, !exclude_no_dm), PatientID))` patients.

* `r nrow(filter(eye_raw_dm, !exclude_no_dm, exclude_no_va))` eyes did not have a visual acuity assessment within 2 weeks prior to or including the index date and were excluded.   

* `r nrow(filter(eye_raw_dm, !exclude_no_dm, !exclude_no_va, exclude_no_thickness))` eyes did not have any OCT retinal thickness measurements (indicative of a usable volume scan) at any point.   

A total of `r nrow(eye_dm)` eyes from `r nrow(distinct(eye_dm, PatientID))` patients remained after these exclusions. 

# Descriptive statistics at baseline

Descriptive statistics for the eyes included in the analysis are given in Tables \@ref(tab:descriptives) to \@ref(tab:va-baseline). Statistics are presented at the eye level. 

## Patient characteristics

```{r descriptives}
bind_rows(
  eye_dm %>%
    transmute(Age = index_age) %>%
    GenerateDescriptives(),
  eye_dm %>%
    transmute(Age = index_age, Gender, Eyes = " ") %>%
    GenerateDescriptives(type = c(Age = "Mean (Range)"))
) %>%
  kbl(caption = "Baseline characteristics of eyes with DM. Frequency and percentage reported for categorical variables.") %>%
  kable_paper()
```

```{r follow-up-length}
eye_dm %>% 
  transmute(Eyes = " ",
            `Years followup` = years_followup,
            Encounters = encounters,
            `OCT thickness measurements` = oct_thickness_n            
            
            ) %>% 
  GenerateDescriptives(type = c(`Years followup` = "Median (Range)", 
                                Encounters = "Median (Range)",
                                `OCT thickness measurements` = "Median (Range)")) %>%
 kbl(caption = "Length of follow-up, number of clinic encounters (both virtual and in-person) and number of OCT thickness measurements.") %>% 
  kable_paper()

```

```{r va-baseline}
bind_rows(
  va_history_dm %>% 
    filter(baseline) %>% 
    transmute(va_logmar) %>% 
    GenerateDescriptives(),
va_history_dm %>% 
  filter(baseline) %>% 
  transmute(va_logmar, va_etdrs, va_category_snellen, Eyes = "n") %>% 
  GenerateDescriptives(type = c(va_logmar = "Median (IQR)", va_etdrs = "Median (IQR)"))) %>% 
  left_join(var_desc, by = "Variable") %>%
  mutate(Description = coalesce(Description, Variable)) %>%
  select(Description, everything(), -Variable) %>%
  kbl(caption = "Distribution of visual acuity outcomes at baseline.") %>% 
  kable_paper()
```
