---
title: "GA study - Cohort construction"
author: "David M Wright - d.wright@qub.ac.uk"
date: "Document compiled: `r Sys.Date()`"
output: 
  bookdown::html_document2:
     toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(kableExtra)
library(rprojroot)
library(tidyverse)
library(broom)
```

`r R.version.string`   

# Introduction

```{r run-analysis, include=FALSE}
source(find_rstudio_root_file("Code/GA-cohort-construction.R"))
```

This analysis is of Electronic Medical Records (EMR) from the Belfast Health and Social Care Trust Macular Service extracted over the period `r min(as.Date(encounters$EncounterDate))` to `r max(as.Date(encounters$ExtractDate))`. The aim of the study is to investigate development of geographic atrophy (GA).

This document describes construction of the analysis cohort. Pre-processing of the Medisoft data is described in `BIRAX-assemble-Medisoft-data.html`.

 
# Cohort construction

Data were extracted on `r as.Date(max(patients$ExtractDate))`. A proportion of event dates were after the recorded `ExtractDate`. For example, in the `Encounters` table `r nrow(filter(encounters, EncounterDate > ExtractDate))` records had an `EncounterDate` after the `ExtractDate`. This is related to the date perturbation used by Medisoft to maintain patient anonymity in the extract.

The following steps were taken to process the data:   

Diagnoses   

* Identification of eyes with a diagnosis of GA (`CoPathologyCode` "R338" in `BELCoPathology` table, or `DiagnosisDescription` containing "geographic atrophy" in `BELDiagnoses` table, or `ClinicalTermDescription` containing "geographic atrophy" in `BELClinicalFindings`).    


## Inclusion and exclusion criteria

Data were extracted by Medisoft with the following inclusion and  exclusion criteria. 

Inclusion: "Any patient having attended the BHSCT macular treatment service at least once during the period May 2008 to May 2018 with a diagnosis of AMD, DR/DMO or RVO."   
Exclusion: "Any patient marked as a test patient."

EMR for a total of `r nrow(patients)` patients were included in the extracted dataset. The unit of analysis is the eye so criteria were applied at the eye level:   

* The raw data potentially contained `r nrow(eye_raw_ga)` eyes (some were not noted in any of the tables and so assumed normal).   
* `r nrow(filter(eye_raw_ga, exclude_no_GA))` eyes had no diagnosis of GA and were excluded.    

This left `r nrow(filter(eye_raw_ga, !exclude_no_GA))` GA eyes from `r nrow(distinct(filter(eye_raw_ga, !exclude_no_GA), PatientID))` patients.

* `r nrow(filter(eye_raw_ga, !exclude_no_GA, exclude_no_va))` eyes did not have a visual acuity assessment within 2 weeks prior to or including the index date and were excluded.   

* `r nrow(filter(eye_raw_ga, !exclude_no_GA, !exclude_no_va, exclude_no_thickness))` eyes did not have any OCT retinal thickness measurements (indicative of a usable volume scan) at any point.   

A total of `r nrow(eye_ga)` eyes from `r nrow(distinct(eye_ga, PatientID))` patients remained after these exclusions. 

# Descriptive statistics at baseline

Descriptive statistics for the eyes included in the analysis are given in Tables \@ref(tab:descriptives) to \@ref(tab:va-baseline). Statistics are presented at the eye level as some patients developed GA in both eyes at different times. 

## Patient characteristics

```{r descriptives}
bind_rows(
  eye_ga %>%
    transmute(Age = index_age) %>%
    GenerateDescriptives(),
  eye_ga %>%
    transmute(Age = index_age, Gender, Eyes = " ") %>%
    GenerateDescriptives(type = c(Age = "Mean (Range)"))
) %>%
  kbl(caption = "Baseline characteristics of eyes with GA. Frequency and percentage reported for categorical variables.") %>%
  kable_paper()
```

```{r follow-up-length}
eye_ga %>% 
  transmute(Eyes = " ",
            `Years followup` = years_followup,
            `Years pre-GA` = years_pre_GA,
            `Years with GA` = years_with_GA,
            Encounters = encounters,
            `OCT thickness measurements` = oct_thickness_n            
            
            ) %>% 
  GenerateDescriptives(type = c(`Years followup` = "Median (Range)", 
                                `Years pre-GA` = "Median (Range)",
                                `Years with GA` = "Median (Range)",
                                Encounters = "Median (Range)",
                                `OCT thickness measurements` = "Median (Range)")) %>%
 kbl(caption = "Length of follow-up, number of clinic encounters (both virtual and in-person) and number of OCT thickness measurements.") %>% 
  kable_paper()

```

```{r va-baseline}
bind_rows(
  va_history_ga %>% 
    filter(baseline) %>% 
    transmute(va_logmar) %>% 
    GenerateDescriptives(),
va_history_ga %>% 
  filter(baseline) %>% 
  transmute(va_logmar, va_etdrs, va_category_snellen, Eyes = "n") %>% 
  GenerateDescriptives(type = c(va_logmar = "Median (IQR)", va_etdrs = "Median (IQR)"))) %>% 
  left_join(var_desc, by = "Variable") %>%
  mutate(Description = coalesce(Description, Variable)) %>%
  select(Description, everything(), -Variable) %>%
  kbl(caption = "Distribution of visual acuity outcomes at baseline.") %>% 
  kable_paper()
```
