---
title: "Rosetrees study - Cohort construction"
author: "David M Wright - d.wright@qub.ac.uk"
date: "Document compiled: `r Sys.Date()`"
output: 
  bookdown::html_document2:
     toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(kableExtra)
library(rprojroot)
library(tidyverse)
library(broom)
```

`r R.version.string`   

# Introduction

```{r run-analysis, include=FALSE}
source(find_rstudio_root_file("Code/RT-cohort-construction.R"))
```

This analysis is of Electronic Medical Records (EMR) from the Belfast Health and Social Care Trust Macular Service extracted over the period `r min(as.Date(encounters$EncounterDate))` to `r max(as.Date(encounters$ExtractDate))`. The aim of the study is to investigate models to predict second eye conversion to wet AMD.

This document describes construction of the analysis cohort. It is similar to the BIRAX study cohort described in `BIRAX-cohort-construction.html`. Pre-processing of the EMR (Medisoft) and fluid data are described in `BIRAX-assemble-Medisoft-data.html` and `BIRAX-assemble-NOA-data.html` respectively.


# Cohort construction

Data were extracted on `r as.Date(max(patients$ExtractDate))`. A proportion of event dates were after the recorded `ExtractDate`. For example, in the `Encounters` table `r nrow(filter(encounters, EncounterDate > ExtractDate))` records had an `EncounterDate` after the `ExtractDate`. This is related to the date perturbation used by Medisoft to maintain patient anonymity in the extract.

Data processing steps are described in `BIRAX-cohort-construction.html`.
Additional steps specific to this analysis were:

* Identification of the first injection recorded for each patient and the date (the index date).   
* Calculated the time between the index date and each OCT visit (potentially negative for OCTs before the index date).   
* Generated an OCT visit sequence for each eye, with visit 0 at the index date, negative values prior to the index date and positive values after.   


## Inclusion and exclusion criteria

Data were extracted by Medisoft with the following inclusion and  exclusion criteria. 

Inclusion: "Any patient having attended the BHSCT macular treatment service at least once during the period May 2008 to May 2018 with a diagnosis of AMD, DR/DMO or RVO."   
Exclusion: "Any patient marked as a test patient."

The unit of analysis is the person so criteria were applied at the person level:   

* The raw data contained `r nrow(patients_raw_rt)` patients.   
* `r nrow(filter(patients_raw_rt, exclude_no_AMD))` patients where neither eye had a diagnosis of age-related macular degeneration during the study period were excluded. 
* `r nrow(filter(patients_raw_rt, !exclude_no_AMD, exclude_RVO))` patients that had a diagnosis of RVO in either eye at any point during the study period were excluded.   
* `r nrow(filter(patients_raw_rt, !exclude_no_AMD, !exclude_RVO, exclude_DR_DMO))` patients that had a diagnosis of DMO/DR in either eye at any point during the study period were excluded.   
* `r nrow(filter(patients_raw_rt, !exclude_no_AMD, !exclude_RVO, !exclude_DR_DMO, exclude_no_injections))` patients had no anti-VEGF injections in either eye during the study period and were excluded.   
* `r nrow(filter(patients_raw_rt, !exclude_no_AMD, !exclude_RVO, !exclude_DR_DMO, !exclude_no_injections, exclude_age))` patients were aged <50 at the index date (first injection in first eye) and were excluded.   
* `r nrow(filter(patients_raw_rt, !exclude_no_AMD, !exclude_RVO, !exclude_DR_DMO, !exclude_no_injections, !exclude_age, exclude_no_va))` patients did not have a visual acuity assessment in both eyes within 2 weeks prior to or including the index date and were excluded.   
* `r nrow(filter(patients_raw_rt, !exclude_no_AMD, !exclude_RVO, !exclude_DR_DMO, !exclude_no_injections, !exclude_age, !exclude_no_va, exclude_no_oct))` patients did not have OCT volume scans taken of both eyes on or up to 2 weeks prior to the index date and were excluded.   
* `r nrow(filter(patients_raw_rt, !exclude_no_AMD, !exclude_RVO, !exclude_DR_DMO, !exclude_no_injections, !exclude_age, !exclude_no_va, !exclude_no_oct, exclude_single_oct))` patients that had no further OCT scans after the initial scan were excluded (not possible to assess the outcome as not being regularly monitored).   

A total of `r nrow(patients_rt)` patients remained after these exclusions. 


# Study period

Follow-up for each patient started at `fup_start`, the date of the OCT scan immediately prior to the first anti-VEGF injection for the first eye (within a maximum of two weeks prior to the injection).  
Follow-up ended at `fup_end`, the date of the first anti-VEGF injection in the second eye (`index_date_2)` if conversion to wet AMD occurred or the date of the last OCT scan (`oct_end`), if conversion did not occur.   
The primary outcome, conversion of the second eye to wet AMD, was defined as a logical variable `conversion`.  
The total follow-up in months was defined in `fup_months`.  


OCT visits outside for the follow-up period for each eye were excluded leaving a total of `r nrow(oct_history_rt)` scans.


# Descriptive statistics at baseline

Descriptive statistics for the eyes included in the analysis are given in Tables \@ref(tab:descriptives) to \@ref(tab:fluid-grid-baseline). Statistics are presented at the eye level as some patients were treated in both eyes, with treatment starting at different times. 

## Patient characteristics

```{r descriptives}
patients_rt %>%
   transmute(Age = index_age, Gender, Patients = " ") %>%
  GenerateDescriptives(type = c(Age = "Mean (Range)")) %>%
  kbl(caption = "Baseline characteristics of patients with AMD. Frequency and percentage reported for categorical variables.") %>%
  kable_paper()
```


```{r eye-baseline}
patients_rt %>% 
  transmute(`Visual acuity (logMAR) - eye 1` = va_logmar_1,
            `Visual acuity (logMAR) - eye 2` = va_logmar_2,
            `Visual acuity (ETDRS) - eye 1` = va_etdrs_1,
            `Visual acuity (ETDRS) - eye 2` = va_etdrs_2,
            `Laterality - eye 1` = EyeCode_1) %>% 
  GenerateDescriptives(type = c(`Visual acuity (ETDRS) - eye 1` = "Median (IQR)",
                                `Visual acuity (ETDRS) - eye 2` = "Median (IQR)")) %>% 
  kbl(caption = "Baseline characteristics of first treated eye and second eye.") %>%
  kable_paper()
```

```{r conversion}
patients_rt %>%
   transmute(conversion = if_else(conversion, "Converted", "Did not convert"), `Follow up (months)` = fup_months, Patients = " ") %>%
  GenerateDescriptives(col_var = conversion, type = c(`Follow up (months)` = "Median (IQR)")) %>%
  kbl(caption = "Length of follow up by second eye AMD conversion status.") %>%
  kable_paper()
```

```{r oct-visits}
oct_history_rt %>% 
  group_by(PatientID,  eye_order) %>% 
  summarise(`OCT sequence length` = n(), .groups = "drop") %>% 
  # mutate()
  select(-PatientID) %>% 
  GenerateDescriptives(col_var = eye_order, type = c(`OCT sequence length` = "Median (Range)")) %>% 
  bind_rows(oct_history_rt %>% 
  transmute(eye_order, "Total OCT visits" = " ") %>% 
  GenerateDescriptives(col_var = eye_order)) %>% 
  kbl(caption = "Distribution of OCT visits by eye order") %>% 
  kable_paper()
```


```{r sequence-sample, fig.cap = "Sample of OCT visit sequences.", fig.height=10}
oct_history_rt %>% 
  slice(1:2000) %>% 
  mutate(Patient = abbreviate(PatientID)) %>% 
  rename(`Months since index date` = months_since_index) %>% 
ggplot(aes(x = `Months since index date`, y = Patient)) + geom_point(aes(colour = conversion)) +
  theme_light()
```
```{r sequence-length, fig.cap = "Distribution of OCT sequence lengths"}
patients_rt %>% 
  rename(`OCT sequence length` = oct_visits) %>% 
  ggplot(aes(x = `OCT sequence length`)) + 
  geom_bar() + 
  theme_light() +
  ylab("Patients")
```

