---
title: "BIRAX study - TASMC descriptive statistics"
author: "David M Wright - d.wright@qub.ac.uk"
date: "Document compiled: `r Sys.Date()`"
output: 
  bookdown::html_document2:
     toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(kableExtra)
library(rprojroot)
library(tidyverse)
library(broom)
library(janitor)
```

`r R.version.string`   

# Introduction

```{r run-analysis, include=FALSE}
source(find_rstudio_root_file("Code/TASMC-descriptive-statistics.R"))
```

This analysis is of Electronic Medical Records (EMR) from the Tel Aviv Sourasky Medical Centre extracted over the period `r min(mdc$Date)` to `r max(mdc$Date)`. The aim of the study is to investigate the potential for machine learning to identify factors and biomarkers associated with long-term AMD treatment response.

This document gives descriptive statistics for various sub-groups in the dataset.

# VA change by baseline visual acuity{#VA-section}

```{r va-change-by-year}
# Need to use follow up years rather than treatment years because the snapshots are only defined for these.
va_history %>% 
  filter(baseline & snapshot) %>% 
  select(PatientID, EyeCode, va_category_snellen) %>% 
  inner_join(
va_history %>% 
  filter(snapshot,
         follow_up_year != 0 , follow_up_year != 0.5) %>% 
  select(PatientID, EyeCode, va_change_logmar, follow_up_year),
by = c("PatientID", "EyeCode")) %>% 
  select(-PatientID, -EyeCode) %>% 
  group_by(va_category_snellen, follow_up_year) %>% 
  skim_numeric() %>% 
  format_skim_numeric(col_var = follow_up_year) %>% 
  filter(Level %in% c("n", "Mean (SD)")) %>% 
  arrange(Level) %>% 
  select(-Variable) %>% 
  rename(`Baseline VA Category` = va_category_snellen) %>% 
  kbl(caption = "Distribution of VA changes (logMAR from baseline) at end of each year of anti-VEGF treatment by baseline VA category.") %>% 
  kable_paper()
```

```{r va-change-category-by-year}
va_history %>% 
  filter(snapshot,
         follow_up_year != 0 , 
         follow_up_year != 0.5,
         !is.na(va_change_cat)) %>% 
  transmute(va_category_snellen, follow_up_year,  va_change_cat, Eyes = "n") %>% 
  group_by(va_category_snellen) %>% 
  nest() %>% 
  mutate(descr = map(.x = data, ~GenerateDescriptives(.x, col_var = follow_up_year))) %>% 
  arrange(va_category_snellen) %>% 
  unnest(descr) %>% 
  select(-Variable, -data) %>% 
  rename(`Baseline VA Category` = va_category_snellen) %>% 
  rename(`Change category` = Level) %>% 
  kbl(caption = "Distribution of VA change categories (from baseline) at the end of each year of anti-VEGF treatment. Ceiling effects not adjusted for.") %>% 
  kable_paper()
```


# Phases of functional response

The response to treatment in terms of visual acuity change can be divided into three phases:

* Primary response - measured at the first visit after the third anti-VEGF injection.   
* Secondary response - measured at the second visit after the third anti-VEGF injection, provided this visit was <12 months since the first injection.   
* Late response - measured at the first visit in the second year since the first injection. 

The distribution of VA measurements across the phases is given in Figure \@ref(fig:phase-time). Changes in treatment response between phases are described in Tables \@ref(tab:first-transition) and \@ref(tab:second-transition). Due to the timing of some of the events, not all eyes can be assigned a complete set of phases, hence the 'NA' columns.


```{r phase-time, fig.cap="Timing of VA measurements used to represent phases of treatment response.", warning = FALSE, fig.height=3}
ggplot(va_response, aes(x = years_since_index)) +
  geom_histogram(bins = 40) +
  facet_wrap(~response) +
  xlim(c(0,2)) +
  theme_light()
```


```{r first-transition}
va_phases %>% 
  select(Primary, Secondary) %>% 
  GenerateDescriptives(col_var = Secondary) %>% 
  filter(!is.na(Level)) %>% 
  select(-Variable) %>% 
  rename(`Primary response` = Level) %>% 
  kbl(caption = "Visual acuity changes relative to baseline in the primary and secondary phases of treatment response. Columns indicate response in the secondary phase.") %>% 
  kable_paper()
```

```{r second-transition}
va_phases %>% 
  select(Secondary, Late) %>% 
  GenerateDescriptives(col_var = Late) %>% 
  filter(!is.na(Level)) %>% 
  select(-Variable) %>% 
  rename(`Secondary response` = Level) %>% 
  kbl(caption = "Visual acuity changes relative to baseline in the secondary and late phases of treatment response. Columns indicate response in the late phase.") %>% 
  kable_paper()
```

# Fluid response by baseline fluid characteristics

The same classification of response phases was used to [this section](#VA-section).

```{r fluid-phase-time, fig.cap="Timing of fluid measurements used to represent phases of treatment response.", warning = FALSE, fig.height=3}
ggplot(fluid_response, aes(x = years_since_index)) +
  geom_histogram(bins = 40) +
  facet_wrap(~response) +
  xlim(c(0,2)) +
  theme_light()
```

```{r fluid-first-transition}
fluid_phases %>% 
  select(Primary, Secondary) %>% 
  GenerateDescriptives(col_var = Secondary) %>% 
  filter(!is.na(Level)) %>% 
    select(-Variable) %>% 
  rename(`Primary response` = Level) %>% 
  kbl(caption = "Fluid changes relative to baseline in the primary and secondary phases of treatment response. Columns indicate response in the secondary phase.") %>% 
  kable_paper()
```

```{r fluid-second-transition}
fluid_phases %>% 
  select(Primary, Secondary) %>% 
  GenerateDescriptives(col_var = Secondary) %>% 
  filter(!is.na(Level)) %>% 
    select(-Variable) %>% 
  rename(`Primary response` = Level) %>% 
  kbl(caption = "Fluid changes relative to baseline in the primary and secondary phases of treatment response. Columns indicate response in the late phase.") %>% 
  kable_paper()
```


```{r fl-response-amd-type}
fluid_history %>% 
  filter(baseline, amd_type != "No fluid") %>% 
  select(PatientID, EyeCode, baseline, amd_type, one_of(noa_fluid)) %>% 
  inner_join(fluid_response %>% 
             select(PatientID, EyeCode, response, fl_response), by = c("PatientID", "EyeCode")) %>% 
  select(-IRF, -SRF, -Fluid, -PatientID) %>% 
  group_by(amd_type, response, fl_response) %>% 
  skim_numeric() %>% 
  format_skim_numeric(col_var = fl_response) %>% 
  # filter(Level == "Median (IQR)" | (Level == "n" & Variable == "SRFVolumeNl")) %>% 
  filter(Level == "Median (IQR)") %>% 
  mutate(Variable = if_else(Level == "n", "Eyes", Variable)) %>% 
  arrange(amd_type, Variable, response, Level) %>% 
  select(Variable, amd_type, response, Level, Good, Partial, `Non-responder`) %>% 
  kbl(caption ="Baseline fluid measurements by response and phase after treatment.") %>% 
  kable_paper()
```


Figures \@ref(fig:first-resolution-injections) and \@ref(fig:first-resolution-time) and Table \@ref(tab:first-resolution-stats) do not include those where fluid never resolved.

```{r first-resolution-injections, fig.cap="Number of injections required to achieve fluid resolution."}
first_resolution %>% 
  ggplot(aes(x = injection_number)) + 
  geom_bar() +
  theme_light()
```
```{r first-resolution-time, fig.cap="Time required to achieve fluid resolution."}
first_resolution %>% 
  ggplot(aes(x = months_since_index)) + 
  geom_histogram(bins = 40) +
  theme_light()
```

```{r first-resolution-stats}
first_resolution %>% 
  transmute(cc ="Value", injection_number, months_since_index) %>% 
  group_by(cc) %>% 
  skim_numeric() %>% 
  format_skim_numeric(col_var = cc) %>% 
  filter(Level %in% c("n", "Median (IQR)", "Mean (SD)")) %>% 
  kbl(caption = "Time and number injections required to achieve fluid resolution.") %>% 
  kable_paper()
```


# Number of annual injections

The number of injections per year was calculated across treatment years 2 and 3 (i.e. total number of injections / 2). This rate was categorised into three groups and used to classify various outcomes.


```{r inj-by-year-baseline-va}
injections_per_yr %>% 
  inner_join(va_history, by = c("PatientID", "EyeCode")) %>% 
  filter(baseline) %>% 
  transmute(injections_per_yr_grp, va_logmar, va_category_snellen, Eyes = "n") %>% 
  GenerateDescriptives(col_var = injections_per_yr_grp, type = c(va_logmar = "Median (IQR)")) %>% 
  kbl(caption = "VA at baseline by number of injections per year in years 2-3 (aggregated).") %>% 
  kable_paper()
```

```{r inj-by-year-baseline-fluid-status}
injections_per_yr %>% 
  inner_join(fluid_history, by = c("PatientID", "EyeCode")) %>% 
  filter(baseline) %>% 
  select(injections_per_yr_grp, amd_type) %>% 
  GenerateDescriptives(col_var = injections_per_yr_grp) %>% 
  select(-Variable) %>% 
  rename(`AMD type` = Level) %>% 
  kbl(caption = "AMD type at baseline (SRF/IRF status) by number of injections per year in years 2-3 (aggregated).") %>% 
  kable_paper()
```


```{r inj-by-year-baseline-fluid}
injections_per_yr %>% 
  inner_join(fluid_history, by = c("PatientID", "EyeCode")) %>% 
  filter(baseline) %>% 
  select(injections_per_yr_grp, one_of(noa_fluid)) %>% 
  select(-IRF, -SRF, -Fluid) %>% 
  group_by(injections_per_yr_grp) %>% 
  skim_numeric() %>% 
  format_skim_numeric(col_var = injections_per_yr_grp) %>% 
  filter(Level == "Median (IQR)" | (Level == "n" & Variable == "SRFVolumeNl")) %>% 
  mutate(Variable = if_else(Level == "n", "Eyes", Variable)) %>% 
  arrange(Level) %>% 
  kbl(caption = "Baseline fluid measurements by number of injections per year in years 2-3 (aggregated).") %>% 
  kable_paper()

```

```{r injections-by-year}
injections_per_yr %>% 
  left_join(injections_yr1, by = c("PatientID", "EyeCode")) %>% 
  select(injections_per_yr_grp, injections_yr1) %>% 
  group_by(injections_per_yr_grp) %>% 
  skim_numeric() %>% 
  format_skim_numeric(col_var = injections_per_yr_grp) %>% 
  filter(Level %in% c("n", "Median (IQR)")) %>% 
  arrange(Level) %>% 
  select(-Variable) %>% 
  mutate(Level = if_else(Level == "n", "Eyes", Level)) %>% 
  kbl(caption = "Number of injections in the first year by number of injections per year in years 2-3 (aggregated).") %>% 
  kable_paper()
```


```{r inj-by-year-response}
injections_per_yr %>% 
  inner_join(va_phases, by = c("PatientID", "EyeCode")) %>% 
  filter(!is.na(Primary)) %>% 
  select(injections_per_yr_grp, Primary) %>% 
  GenerateDescriptives(col_var = injections_per_yr_grp) %>% 
  select(-Variable) %>% 
  rename(`Primary response` = Level) %>% 
  kbl(caption = "Initial treatment response (primary phase) by number of injections per year in years 2-3 (aggregated).") %>% 
  kable_paper()
```

